---
title: "Intro R Workshop: Exercise 2"
author:
  - affiliation: University of the Western Cape
  - name: AJ Smit
date: "17 March 2022"
output:
  rmarkdown::pdf_document:
    toc: no
    dev: "cairo_pdf"
    df_print: tibble
    fig_caption: yes
    highlight: default
    keep_tex: yes
    latex_engine: xelatex
    template: /Volumes/GoogleDrive/My Drive/repos/workshops/Intro_R_Workshop/_GitBook/template-1.tex
papersize: A4
fontsize: 10pt
geometry: margin=1in
tables: yes
language: Australian
mainfont: Minion Pro
sansfont: Myriad Pro
monofont: Courier New
csl: /Volumes/GoogleDrive/My Drive/repos/workshops/Intro_R_Workshop/_GitBook/ecography.csl
bibliography: /Volumes/GoogleDrive/My Drive/repos/bibliography/bibliography.bib
editor_options: 
  chunk_output_type: inline
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(include = FALSE, echo = FALSE, warning = FALSE,
                      message = FALSE, tidy = FALSE, cache = FALSE,
                      fig.align = 'left', fig.path = 'figures/exercise-1')
```

```{r load-pkg, include = FALSE}
library(RmarineHeatWaves)
library(data.table)
library(plyr)
library(tidyverse)
library(lubridate)
library(forcats)
library(mgcv)
library(broom)
library(readr)
library(doMC); doMC::registerDoMC(cores = 4)
library(ggpubr)
```

# The SACTNmonthly_v4.0.RData

Please *exactly* recreate the figure immediately below (you may use your own colour for the line). Note: in order to calculate a yearly mean for each of the data points within a year, you will have to use one of the functions in the **lubridate** package. There is also the `mutate()` function (within the **dplyr** package) that I have mentioned before, but which we have not explicitely practiced --- it will have to be used to receive the result of the **lubridate** function that I alluded to above.

```{r ex-1, echo = FALSE, include = TRUE, fig.width = 5, fig.height = 6, fig.align = 'center'}
load("../_GitBook/data/SACTNmonthly_v4.0.RData")
dat1 <- as_tibble(SACTNmonthly_v4.0)
rm(SACTNmonthly_v4.0)

dat2 <- dat1 %>% 
  mutate(date.yr = year(date)) %>% 
  group_by(site, src, date.yr) %>% 
  summarise(mn.temp = mean(temp, na.rm = TRUE),
            sd.temp = sd(temp, na.rm = TRUE)) %>% 
  ungroup() %>% 
  filter(src == "KZNSB")

ggplot(data = dat2, aes(x = date.yr, y = mn.temp)) +
  geom_line(aes(group = site), colour = "salmon") +
  facet_wrap(~site, ncol = 5) +
  scale_x_continuous(breaks = c(1980, 2000)) +
  scale_y_continuous(breaks = c(20, 22, 24)) +
  labs(x = "Year", y = "Temperature (°C)", title = "KZNSB: series of annual means") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```

# The data.laminaria.csv data

Please recreate the following figure *exactly* (note: the graph is not atually meaningful, and it is the incorrect way to display the data; used here for demonstration purpuses only):

```{r ex-2, include = TRUE, echo = FALSE, fig.height = 3}
dat3 <- read_csv(file = "../_GitBook/data/laminaria.csv") %>% 
  filter(region == "FB")

ggplot(dat3, aes(x = blade_length, y = blade_weight)) +
  geom_point(aes(colour = site)) +
  geom_line(aes(group = site, colour = site)) +
  scale_colour_brewer(palette = "Accent") +
  facet_wrap(~site) +
  labs(x = "Blade length (cm)", y = "Blade mass (kg)", title = "A crazy graph of some data for False Bay sites") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```

In the above graph I used one of the palettes included with the Colour Brewer scale. Unfortunately the plot for Romans Rock is now missing. Why? Please provide a solution to this problem --- i.e. make a new graph where the problem is no longer present. Combine the graphs as two sub-plots (i.e. a figure labeled 'A' and 'B') using the facility offered by the `ggpubr` package.

# The ToothGrowth data

These data reside in `datasets::ToothGrowth`. Please produce a graph like the one below. The adjustment of the error bars (here showing ±SD) is a bit tricky, so you will have to figure out how to consult the help files, or find alternative help somewhere using an internet search.

```{r ex-3, echo = FALSE, include = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
dat4 <- datasets::ToothGrowth

dat5 <- dat4 %>% 
  group_by(supp, dose) %>% 
  summarise(mn.len = mean(len),
            sd.len = sd(len))

ggplot(data = dat5, aes(x = dose, y = mn.len)) +
  geom_col(aes(fill = supp), colour = "black", position = "dodge") + 
  geom_errorbar(aes(ymin = mn.len - sd.len, ymax = mn.len + sd.len, group = supp), 
                position = position_dodge(width = 0.45), width = 0.2) +
  labs(x = "Dose (mg/d)", y = "Tooth length (mm)") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```

```{r}
ggplot(data = dat5, aes(x = dose, y = mn.len)) +
  geom_col(aes(fill = supp), colour = "black", position = position_dodge(width = 0.47)) + 
  geom_errorbar(aes(ymin = mn.len - sd.len, ymax = mn.len + sd.len, group = supp), 
                position = position_dodge(width = 0.47), width = 0.2) +
  labs(x = "Dose (mg/d)", y = "Tooth length (mm)") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```
