---
title: 'Intro R Workshop: Exercise 2'
author:
- affiliation: University of the Western Cape
- name: AJ Smit
date: "17 March 2021"
output:
  html_document:
    df_print: paged
fontsize: 10pt
geometry: margin=1in
tables: yes
language: Australian
mainfont: Minion Pro
sansfont: Myriad Pro
monofont: Courier New
editor_options:
  chunk_output_type: inline
papersize: A4
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(include = FALSE, echo = FALSE, warning = FALSE,
                      message = FALSE, tidy = FALSE, cache = FALSE,
                      fig.align = 'left', fig.path = 'figures/exercise-1')
```

```{r load-pkg, include = FALSE}
library(RmarineHeatWaves)
library(data.table)
library(plyr)
library(tidyverse)
library(lubridate)
library(forcats)
library(mgcv)
library(broom)
library(readr)
library(doMC); doMC::registerDoMC(cores = 4)
library(ggpubr)
```

# 1. The SACTNmonthly_v4.0.RData (A)

Please *exactly* recreate the figure immediately below (you may use your own colour for the line). Note: in order to calculate a yearly mean for each of the data points within a year, you will have to use one of the functions in the **lubridate** package. There is also the `mutate()` function (within the **dplyr** package) that I have mentioned before, but which we have not explicitely practiced --- it will have to be used to receive the result of the **lubridate** function that I alluded to above.

```{r ex-1, echo = FALSE, include = TRUE, fig.width = 5, fig.height = 6, fig.align = 'center'}
load("../_GitBook/data/SACTNmonthly_v4.0.RData")
dat1 <- as_tibble(SACTNmonthly_v4.0)
rm(SACTNmonthly_v4.0)

dat2 <- dat1 %>% 
  mutate(date.yr = year(date)) %>% 
  group_by(site, src, date.yr) %>% 
  summarise(mn.temp = mean(temp, na.rm = TRUE),
            sd.temp = sd(temp, na.rm = TRUE)) %>% 
  ungroup() %>% 
  filter(src == "KZNSB")

ggplot(data = dat2, aes(x = date.yr, y = mn.temp)) +
  geom_line(aes(group = site), colour = "salmon") +
  facet_wrap(~site, ncol = 5) +
  scale_x_continuous(breaks = c(1980, 2000)) +
  scale_y_continuous(breaks = c(20, 22, 24)) +
  labs(x = "Year", y = "Temperature (°C)", title = "KZNSB: series of annual means") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```

# 2. The SACTNmonthly_v4.0.RData (B)

Please produce a series of graphs arranged in the same manner as in the above figure, "KZNSB: series of annual means,", but instead of plotting the time series along the x-axis, produce a monthly climatology (i.e. as per the MS Excel exercise we did on Monday).

# 3. The ToothGrowth data (A)

These data reside in `datasets::ToothGrowth`. Please produce line graphs showing the relationship of tooth length as a function of dose, and ensure that the two types of supplements are adequately distinguished by colour.

# 4. The ToothGrowth data (B)

These data reside in `datasets::ToothGrowth`. Please produce box-and-whisker graphs showing the relationship of tooth length as a function of dose, and ensure that each of the two types of supplements are presented in separate panels (facets).

# 5. The ToothGrowth data (C)

These data reside in `datasets::ToothGrowth`. Please produce a graph like the one below. The adjustment of the error bars (here showing ±SD) is a bit tricky, so you will have to figure out how to consult the help files, or find alternative help somewhere using an internet search.

```{r ex-3, echo = FALSE, include = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
dat4 <- datasets::ToothGrowth

dat5 <- dat4 %>% 
  group_by(supp, dose) %>% 
  summarise(mn.len = mean(len),
            sd.len = sd(len))

ggplot(data = dat5, aes(x = dose, y = mn.len)) +
  geom_col(aes(fill = supp), colour = "black",
           position = position_dodge(width = 0.45)) + 
  geom_errorbar(aes(ymin = mn.len - sd.len, ymax = mn.len + sd.len, group = supp), 
                position = position_dodge(width = 0.45), width = 0.2) +
  labs(x = "Dose (mg/d)", y = "Tooth length (mm)") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```
