---
title: "Animating Angry Ants"
author: "Robert W Schlegel"
date: '`r Sys.Date()`'
output:
  html_document: default
language: Australian
linkcolor: cyan
citecolor: green
urlcolor: cyan
---

```{r animate-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE
)
```

# Introduction
In this supplemenal tutorial we are going to look at how to create animations in R. To do this will require the installation of software outside of R. This software is `ImageMagick` and may be downloaded here: <https://www.imagemagick.org/script/download.php>. Once this software has been installed on your computer it will be necessary to install the `animation` library.

```{r, warning=FALSE, message=FALSE}
# The libraries required for this tut
library(tidyverse)
library(grid)
library(gridExtra)
# install.packages("animation")
library(animation)
```

# Functions for creating ant walks

```{r}
# Calculate speed based on u and v vectors
ant.speed <- function(df){
  df$x2 <- c(NA,df$x[2:nrow(df)] - df$x[1:(nrow(df)-1)])
  df$y2 <- c(NA,df$y[2:nrow(df)] - df$y[1:(nrow(df)-1)])
  speed_abs <- round(sqrt(df$x2^2 + df$y2^2),2)
  speed_abs[is.na(speed_abs)] <- 0
  return(speed_abs)
}

# Create a dataframe with desired number of ants and steps
ant.walk <- function(i,n){
  # Create the random walks
  walk_x <- c(0,round(cumsum(rnorm(n = n-1, mean = 0, sd = 1)),2))
  for(i in 2:i){
  x <- c(0,round(cumsum(rnorm(n = n-1, mean = 0, sd = 1)),2))
  walk_x <- c(walk_x, x)
  }
  walk_y <- c(0,round(cumsum(rnorm(n = n-1, mean = 0, sd = 1)),2))
  for(i in 2:i){
  y <- c(0,round(cumsum(rnorm(n = n-1, mean = 0, sd = 1)),2))
  walk_y <- c(walk_y, y)
  }
  # Create the walking dataframe
  walker <- data.frame(x = walk_x, y = walk_y, 
                       ant = as.factor(rep(1:i, each = n)), 
                       step =  rep(seq(1,n), i))
  walker$speed <- ant.speed(walker)
  walker$speed[walker$step == 1] <- 0
  return(walker)
}
```

# Generate the ants

```{r}
ants <- ant.walk(5, 100)
```

# The function to animate the walk plot

```{r, message=FALSE, warning=FALSE}
walk.plot <- function(i){
  # Map figure
  walk_map <- ggplot(data = ants[ants$step %in% 1:i,], aes(x = x, y = y)) +
    geom_path(aes( group = ant), colour = "gray60") +
    geom_point(data = ants[ants$step == i,], aes(colour = ant))
  # Speed histogram
  walk_hist <- ggplot(data = ants[ants$step %in% 1:i,], aes(x = speed)) +
    geom_histogram() +
    labs(x = "speed")
  # Speed line graph
  walk_line <- ggplot(data = ants[ants$step %in% 1:i,], aes(x = step, y = speed)) +
    geom_line(aes(colour = ant))
  # Wack it together
  grid.arrange(walk_map, walk_hist, walk_line, layout_matrix = cbind(c(1,1), c(1,1), c(2,3)))
}


## Create animation of ts plots
animate.walk.plot <- function() {
  lapply(seq(1,100), function(i) {
    walk.plot(i)
  })
}
```

# Render the GIF

```{r, eval=FALSE}
# By default 'saveGIF()' outputs to the same folder 
# the script where the code is being run from is located
# I have included commented out lines of code here that 
# may be changed to set thedestination for saving the output
# setwd("~/Intro_R_Workshop/BONUS/")
system.time(saveGIF(animate.walk.plot(), interval = 0.2, 
                    ani.width = 800, movie.name = "ant_walk.gif")) ## ~60 seconds
# setwd("~/Intro_R_Workshop/")
```

