---
title: "GIFs"
author: "Robert W Schlegel"
date: '`r Sys.Date()`'
output:
  html_document: default
language: Australian
linkcolor: cyan
citecolor: green
urlcolor: cyan
---

```{r dates-opts, echo=FALSE}
knitr::opts_chunk$set(
  comment = "R>", 
  warning = FALSE, 
  message = FALSE
)
```

# Introduction

Yes my friends, it is true. We may add GIFs to our figures and maps. Rejoice. Better yet, the process is relatively straight forward. We will begin, as usual, by loading our libraries and files.

```{r gif-load}
# Load libraries
library(tidyverse)
library(magick)

# The base image
background <- image_read("../data/air_panel.png") %>% # Load file
  image_scale("900") # Change resolution

# The gif to overlay
anim_overlay <- image_read("../data/carlton_dance.gif")  %>% # Load file 
  image_scale("300") # Change resolution
```

# GIF creation

Once we have loaded our base image and the GIF we want to put on top of it we need to create a function to make these two different file types 'kiss'. With the appropriately named **`magick`** package this is startlingly easy to do.

```{r gif-func}
frames <- lapply(anim_overlay, function(frame) {
  image_composite(background, frame, offset = "+300")
})
```

# GIF animation

With our function for creating the GIF sorted, it is now time to animate it!

```{r gif-anim}
animation <- image_animate(image_join(frames), fps = 10) # FPS = 10 is native speed
```

# GIF save

Jip. Simple as that.

```{r gif-save, eval=FALSE}
image_write(animation, "../BONUS/carlton.gif")
```
