---
title: "Intro R exercises"
subtitle: "Miscellaneous exercises for the mind (2)"  
author: 
  - "AJ Smit"
institute: "University of the Western Cape"
date: '2022/02/18 (updated: `r Sys.Date()`)'
output:
  xaringan::moon_reader:
    css:
      - default
      - "../private/default-fonts.css"
      - ki
      - "../private/uwc_logo.css"
    yolo: false
    self_contained: TRUE  
    lib_dir: libs
    nature:
      highlightStyle: idea
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE, echo=FALSE} 
library(tidyverse)
library(lubridate)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE,
                      comment = "R>",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      fig.retina = 3,
                      fig.width = 6,
                      fig.height = 3.6,
                      fig.align = "center")
```

## Exercise A

**The `shells.csv` data**

* Produce a *tidy* dataset from the data contained in `shells.csv`.
* For each species, relate two measurement variables within the dataset to one-another and represent the relationship with a straight line.
* For each species, concisely produce histograms for each of the measurement variables.
* Use the **colorspace** package and assign interesting colours to your graphs (all graphs above).
* Use the **ggthemr** package and assign interesting themes to your graphs (all graphs above).

---
## Exercise B

**Head Dimensions in Brothers**

Package **boot**, dataset `frets`: The data consist of measurements of the length and breadth of the heads of pairs of adult brothers in 25 randomly sampled families. All measurements are expressed in millimetres.

Please consult the dataset's help file (i.e., load the package **boot** package and type `?frets` on the command line).

* Create a *tidy* dataset from the `frets` data.
* Demonstrate the most concise way for displaying both brother's data on one set of axes.
* Apply your own unique theme modification to the graph in order to produce a publication-worthy figure.

---
## Exercise C

**Results from an Experiment on Plant Growth**

Dataset `PlantGrowth`: Results from an experiment to compare yields (as measured by dried weight of plants) obtained under a control and two different treatment conditions.

* Concisely present the results of the plant growth experiment as graphs:
    - a scatterplot with individual `weight` datapoints as a function of `group`
    - a box and whisker plot showing each `group` (on one set of axes)
    - a bar plot with associated SD for each `group` (on one set of axes)
---
## Exercise D

**Student's Sleep Data**

Dataset `sleep`: Data which show the effect of two soporific drugs (increase in hours of sleep compared to control) on 10 patients.

* Graphically display these data in two different ways.

---
## Exercise E

**English Narrative for Some Code**

Provide an English description for what the following lines of code does.

### Listing 1:

```{r, eval=FALSE}
the_data <- some_data %>%
  mutate(yr = year(date),
         mo = month(date)) %>% 
  group_by(country, yr) %>% 
  summarise(med_chl = mean(chl, na.rm = TRUE)) %>% 
  ungroup()
```

```{r, eval=FALSE}
ggplot(the_data, aes(x = yr, y = med_chl)) +
  geom_line(aes(group = country), colour = "blue3") +
  facet_wrap(~country, nrow = 3) +
  labs(x = "Year", y = "Chlorophyll-a (mg/m3)",
       title = "Chlorophyll-a concentration")
```

---
### Listing 2:

```{r, eval=FALSE}
library(ggforce)
ggplot(iris, aes(Petal.Length, Petal.Width, colour = Species)) +
    geom_point() +
    facet_zoom(x = Species == "versicolor")
```

---
### Listing 3:

```{r, eval=FALSE}
set.seed(13)
my_data = data.frame(
        gender = factor(rep(c("F", "M"), each=200)),
        length = c(rnorm(200, 55), rnorm(200, 58)))
head(my_data)

ggplot(my_data, aes(x = gender, y = length)) +
  geom_boxplot(aes(fill = gender))

ggplot(my_data, aes(x = gender, y = length)) +
  geom_violin()

ggplot(my_data, aes(x = gender, y = length)) +
  geom_dotplot(stackdir = "center", binaxis = "y", dotsize = 0.5)
```