---
title: "Intro R exercises"
subtitle: "Miscellaneous exercises for the mind"  
author: 
  - "AJ Smit"
institute: "University of the Western Cape"
date: '2021/04/21 (updated: `r Sys.Date()`)'
output:
  xaringan::moon_reader:
    css:
      - default
      - "../private/default-fonts.css"
      - ki
      - "../private/uwc_logo.css"
    yolo: false
    self_contained: TRUE  
    lib_dir: libs
    nature:
      highlightStyle: idea
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE, echo=FALSE} 
library(tidyverse)
library(lubridate)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE,
                      comment = "R>",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      fig.retina = 3,
                      fig.width = 6,
                      fig.height = 3.6,
                      fig.align = "center")
```

## Exercise A

---
**Task: the *Laminaria* data**

With all of this information in hand, create a new plot displaying an interesting relationship in the *Laminaria* data.

* Relate two measurement variables within the dataset to one-another
* Use a grouping variable to form a sensible grouping structure
* Make use of two of the geometries `geom_point()`, `geom_line()`, and `geom_smooth()`
* Use one new **ggplot2** theme that you have not yet used

Explain the relationship represented in your figure.

---
## Exercise B

---
- For this exercise, you'll be expected to accomplish Tasks 1, 2 and 3 before producing the final output in Task 4.
- Assemble a properly documented R script in the way you have been taught.

**Task 1: the `datasets::ChickWeight` data**

- Consult the **ggplot2** book that was emailed to you, and construct an interesting graph that addresses some question you are asking about these data
  - use any of the available 'geoms';
  - creatively modify the graph's appearance (but remain sensible and be congnisant of which aesthetics are suitable for publications!);
  - add custom colours, demonstrating some of the colour-handling possibilities that you were taught (or even ones that may not yet have been taught);
  - add the *x*- and *y*-axis titles as appropriate.

---
**Task 2: the `SACTNmonthly_v4.0.Rdata` data**

- Devise any faceted plot showing some aspect of the data you are interested in
  - whatever you choose to graph, please ensure that it results in four facets;
  - apply the same creative process w.r.t. colours, geoms as with Task 1;
  - demonstrate your ability to implement a new aspect of plotting data as per the **ggplot2** book;

---
**Task 3: the `ecklonia.csv` data**

- As above, create two additional single-panel graphs showing aspects not shown in Tasks 1 and 2.

---
**Task 4: the last steps**

- Assemble all four graphs into a 2 x 2 layout using the **ggpubr** functions, as shown earlier.
- Save the R script, and produce a .png file of the final figure.

---
## Exercise C

---
**Task 1: the `SACTNmonthly_v4.0.RData` data, again**

Please *exactly* recreate the figure on the next slide (you may use your own colour for the line). Note: in order to calculate a yearly mean for each of the data points within a year, you will have to use one of the functions in the **lubridate** package. There is also the `mutate()` function (within the **dplyr** package, which comes installed with **tidyverse**) that I have mentioned before, but which we have not explicitely practiced --- it will have to be used to receive the result of the **lubridate** function that I alluded to above.

---
```{r ex-1, echo = FALSE, include = TRUE, fig.width = 5, fig.height = 6, fig.align = 'center'}
load("../_GitBook/data/SACTNmonthly_v4.0.RData")
dat1 <- as_tibble(SACTNmonthly_v4.0)
rm(SACTNmonthly_v4.0)

dat2 <- dat1 %>% 
  mutate(date.yr = year(date)) %>% 
  group_by(site, src, date.yr) %>% 
  summarise(mn.temp = mean(temp, na.rm = TRUE),
            sd.temp = sd(temp, na.rm = TRUE)) %>% 
  ungroup() %>% 
  filter(src == "KZNSB")

ggplot(data = dat2, aes(x = date.yr, y = mn.temp)) +
  geom_line(aes(group = site), colour = "salmon") +
  facet_wrap(~site, ncol = 5) +
  scale_x_continuous(breaks = c(1980, 2000)) +
  scale_y_continuous(breaks = c(20, 22, 24)) +
  labs(x = "Year", y = "Temperature (°C)", title = "KZNSB: series of annual means") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```

---
**Task 2: the `SACTNmonthly_v4.0.RData` data, continue**

Please produce a series of graphs arranged in the same manner as in the above figure, "KZNSB: series of annual means,", but instead of plotting the time series along the x-axis, produce a monthly climatology (i.e. as per the MS Excel exercise we did on Monday).

---
**Task 3. the `datasets::ToothGrowth` data**

These data reside in `datasets::ToothGrowth`. Please produce line graphs showing the relationship of tooth length as a function of dose, and ensure that the two types of supplements are adequately distinguished by colour.

---
**Task 4. the `datasets::ToothGrowth` data, continue**

Produce box-and-whisker graphs showing the relationship of tooth length as a function of dose, and ensure that each of the two types of supplements are presented in separate panels (facets).

---
**Task 5. the `datasets::ToothGrowth` data, continue**

Produce a graph like the one below. The adjustment of the error bars (here showing ±SD) is a bit tricky, so you will have to figure out how to consult the help files, or find alternative help somewhere using an internet search.

```{r ex-3, echo = FALSE, include = TRUE, fig.width = 4, fig.height = 3, fig.align = 'center'}
dat4 <- datasets::ToothGrowth

dat5 <- dat4 %>% 
  group_by(supp, dose) %>% 
  summarise(mn.len = mean(len),
            sd.len = sd(len))

ggplot(data = dat5, aes(x = dose, y = mn.len)) +
  geom_col(aes(fill = supp), colour = "black",
           position = position_dodge(width = 0.45)) + 
  geom_errorbar(aes(ymin = mn.len - sd.len, ymax = mn.len + sd.len, group = supp), 
                position = position_dodge(width = 0.45), width = 0.2) +
  labs(x = "Dose (mg/d)", y = "Tooth length (mm)") +
  theme(panel.grid.major = element_line(size = 0.2, linetype = 2),
        panel.grid.minor = element_line(colour = NA))
```
