---
title: "R datasets"
subtitle: "Built-in datasets"  
author: 
  - "AJ Smit"
date: '2020/07/15 (updated: `r Sys.Date()`)'
output:
  xaringan::moon_reader:
    css:
      - default
      - "/Users/ajsmit/Dropbox/R/workshops/Basic_stats/private/default-fonts.css"
      - ki
      - "/Users/ajsmit/Dropbox/R/workshops/Basic_stats/private/uwc_logo.css"
    yolo: false
    self_contained: TRUE  
    lib_dir: libs
    nature:
      highlightStyle: idea
      highlightLines: true
      highlightLanguage: r
      countIncrementalSlides: false
      ratio: 16:9
---

```{r setup, include=FALSE, echo=FALSE} 
library(tidyverse)
library(ggpubr)
knitr::opts_chunk$set(echo = TRUE,
                      comment = "R>",
                      message = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      fig.retina = 3,
                      fig.width = 6,
                      fig.height = 3.6,
                      fig.align = "center")
```

## Loading and discovering built-in datasets

To find out which datasets are available to you on your system, execute the following command:

```{r}
data()
```

A new tab will open and the list of datasets will be displayed. For example, the following might be displayed:

```
Data sets in package ‘datasets’:

AirPassengers                       Monthly Airline Passenger Numbers 1949-1960
BJsales                             Sales Data with Leading Indicator
BOD                                 Biochemical Oxygen Demand
CO2                                 Carbon Dioxide Uptake in Grass Plants
ChickWeight                         Weight versus age of chicks on different diets
DNase                               Elisa assay of DNase
EuStockMarkets                      Daily Closing Prices of Major European Stock Indices, 1991-1998
Formaldehyde                        Determination of Formaldehyde
HairEyeColor                        Hair and Eye Color of Statistics Students
Harman23.cor                        Harman Example 2.3
[...]
```

---
The exact list might vary from user to user, as this depends on which packages have been installed on your system. For example, scrolling all the way to the bottom of the list produced above using the `data()` command, you'll see the following information and code that can be executed to show all the datasets in all the packages available on your system:

```
Use ‘data(package = .packages(all.available = TRUE))’
```

A particular dataset can be loaded as follow:

```{r, eval=FALSE}
data(acacia, package = "ade4") # the package name is listed also in the data() display
```

You are welcome to use any of these datasets to practice the R commands that you have been taught.
